---
title: "Cricket"
format: 
  html:
    code-fold: true
message: false
warning: false
---
old ipl project. Lots of suboptimal decisions
date:2022-04-17
rendered: 2025-03-31

```{r Load libraries}
# Load necessary libraries
# Using suppressPackageStartupMessages to keep output clean
suppressPackageStartupMessages({
  library(ggtext)
  library(lme4)
  library(tidyverse)
  library(plotly)
  library(viridisLite) # For better color scales
})

```

# Indian Premier League

The goal of the present project is to download, clean, wrangle, and obtain insights from ipl data.

The data will be downloaded from cricsheet.org in a zip format. We will unzip the files to a directory.

```{r Download and unzip files}
#| eval: false
temp <- tempfile()
download.file("https://cricsheet.org/downloads/ipl_csv2.zip", temp)
unzip(zipfile = temp, exdir = "ipl_data") #takes a couple of minutes
rm(temp)
```

The next step would be load the data to R. We can achieve that using the `read_csv` function. However, while this data file is large and contains ball-by-ball details, it does not have match info i.e., who won the match and by how many wickets/runs.

We will obtain that data using the match info files. We can read only those files by adding them as a list for now.

```{r Load match data and info files list}
#| eval: false
df <- read_csv("ipl_data/all_matches.csv") # read the data
files <- list.files(path = "ipl_data", pattern = "*_info.csv") #load the paths of
#metadata files ----
files <- paste0("ipl_data/", files) #append
```

One issue is that there are hundreds of files that match our pattern, and it is not possible to load them and use them effectively if we do it manually. One solution is to write a function that takes each of the file, appends it to a dataframe.

```{r Clean up the info data files and makes them into one file}
#| eval: false
f1 <- function(x) {
  y <- read.table(files[x], #load file
                  sep = ",",
                  skip = 2,
                  col.names = paste0("V",seq_len(5)), fill = TRUE)
  y <- y[,-1] #remove the first column 'info'

  y <- y %>% group_by(V2) %>% #number the column so that it can be pivoted wo errors
    mutate(id = seq_along(V2)) %>%
    ungroup() %>%
    mutate(V2 = paste0(.$V2, .$id)) %>%
    select(1, 2)

  y <- pivot_wider(y, names_from = V2, values_from = V3) %>% # lay data on side
    mutate(match_id = parse_number(file.path(files[x]))) %>%
    select(!starts_with(c("player","regis")))
  # the info file doesn't contain the match id! Parsing it from file names
}

df_info <- f1(1) # read the (meta)data file one
df_info <- df_info[FALSE, ] #keep only the headers
for (i in 1:length(files)) {
  X <- f1(i)

  df_info <- bind_rows(df_info, X)

}

df_info <- df_info %>% 
  rename_with(~str_remove(., "1")) %>%
  select(match_id, match_number, everything()) %>%
  select(!c(event, gender, venue)) %>%
  arrange(match_id)

```

Let's take a look at the data frames we have so far.

```{r Look at dataframes head}
#| eval: false
head(df)
head(df_info)
```

```{r Join dataframes and write CSV}
#| eval: false
final_df <- inner_join(df, df_info, by = 'match_id') # ugly but works..

#Clean up seasons
final_df <- final_df %>% 
    mutate(season.x = case_when(
        season.y == "2007/08" ~ '2008',
        season.y == "2009/10" ~ '2010',
        season.y == "2020/21" ~ '2020',
        TRUE ~ season.y),
        season.x = as.numeric(as.character(season.x))) 

#write.csv(final_df, "final_df.csv")
nanoparquet::write_parquet(final_df, "cricket/final_df.parquet") # 2025 change: saved 100x disk space
```


## Load the final cleaned up data
```{r}
#final_df <- read.csv("final_df.csv")
final_df <- nanoparquet::read_parquet("cricket/final_df.parquet")
```

List of all teams that have played IPL so far
```{r}
(team_list <- final_df %>%
  distinct(team))
```


## Exploring the Data
Lets start by making a function that gives some basic details about any given team
```{r}
team_stat <- function(team_name) {

f1 <- final_df %>% #_team_select
    group_by(season.x) %>%
    arrange(desc(match_id)) %>%
    filter(grepl(team_name, team, ignore.case = T) |
             grepl(team_name, team2, ignore.case = T))

f2 <- distinct(.data = f1, match_id) %>% count() #match count

f3 <- filter(f1, grepl(team_name, winner, ignore.case = T)) %>%#winning matches
    distinct(match_id) %>% count()

f4 <- filter(f1, grepl(team_name, winner, ignore.case = T))

df <- inner_join(f2, f3, by = "season.x") %>%
  rename(matches = n.x, wins = n.y) %>%
  mutate(win_rate = round(wins/matches * 100))


df$team <- f4[1,36] # add a new column with the team name
df <- df %>% unnest(last_col()) %>% 
  rename(team = last_col()) %>% #rename the column
  ungroup() #ungroup the df

return(df)

}
```
This particular function takes regex input

```{r}
team_stat("hyd")

```

```{r}
team_stat("delhi")

team_stat("giants") #vs
team_stat("giant$")
```

## How did any single team perform over the years?
```{r}
plot_team <- function(x3){
p1 <- team_stat(x3)
ggplot(p1, aes(x = as.factor(season.x), y = win_rate, group = 1)) +
  geom_point() +
  geom_line() +
  scale_y_continuous(limits = c(0, 100))+
  labs(
    x = "Season",
    y = "Win Rate",
    title = paste(first(p1[1,5]), "over the years")) #grabs the name of the team
}
```

Performance of SRH(DC) over the years
```{r}
plot_team("hyd|deccan") %>% ggplotly()
```

Delhi
```{r}
plot_team("delhi") %>% ggplotly()
```

Performance of all the teams over the years
```{r Plot of winning rates, message=FALSE, warning=FALSE}
team_list2 <- team_list %>% filter(!grepl("X", team)) # Remove KXIPunjab

d2 <- NULL # create a empty df for stats of all teams
for (i in 1:nrow(team_list2)) {
  d1 <- team_stat(word(team_list2$team[i], 1))
  d2 <- bind_rows(d2,d1)
}

p4 <- ggplot(d2, aes(x = season.x, y = win_rate, group = 1)) +
  geom_point() +
  geom_line() +
  facet_wrap(~team)+
  scale_y_continuous(limits = c(0, 100))+
  labs(
    x = "",
    y = "Win Rate",
    title = "Winning Rate of the teams over the years")

ggplotly(p4)

```


Which team has won the most matches?

```{r}
final_df %>%
  distinct(match_id,.keep_all = TRUE) %>%
  count(winner) %>%
  arrange(desc(n)) %>%
  head(5) %>%
  ggplot(aes(x = reorder(winner,desc(n)), y = n))+
  geom_col()+
  labs(
    x = "Team",
    y = "Number of Wins")+
  scale_x_discrete(label = function(x) str_trunc(x, 19))+
  scale_y_continuous(expand = c(0,0), limits= c(0, 200))

```

How many seasons has the IPL been going on for?
```{r}
ipl_season <- final_df %>%  
   distinct(season.x)
ipl_season %>% count()
```

Lets write a function to get the top-scoring batmen of any given year
```{r}
high_scorer <- function(x = "", y = "", n = 10){
# x is a specific year
# y is a specific player
# n is the number of results to show
  y1 <- final_df %>%
    filter(grepl(x, season.x) & grepl(y, striker, ignore.case = T)) %>%
    group_by(striker) %>%
    summarise(runs = sum(runs_off_bat)) %>%
    arrange(desc(runs)) %>%
    head(n)


  y1 <- y1 %>% mutate(rank = row_number(), year = x) %>%
    select(3,2,1,4) %>% ungroup()

  #write a loop to generate boundaries, averages etc. for each batsmen?

  bound <- NULL
  for (i in min(y1$rank):max(y1$rank)) {
    striker_interest <- y1$striker[i]

    df_interest <- final_df %>% filter(grepl(x, season.x) &
                                         striker %in% striker_interest)

    sixes <- df_interest %>%
      group_by(striker) %>%
      filter(runs_off_bat == 6) %>%
      count() %>%
      ungroup() %>%
      rename(sixes = n)

    fours <- df_interest %>%
      group_by(striker) %>%
      filter(runs_off_bat == 4) %>%
      count() %>%
      ungroup() %>%
      rename(fours = n)

    fifties <- df_interest %>%
      group_by(striker, match_id) %>%
      summarise(runs = sum(runs_off_bat)) %>%
      filter(runs >= 50) %>% count() %>% ungroup() %>%
      rename(fifties = n)

    hundreds <- df_interest %>%
      group_by(striker, match_id) %>%
      summarise(runs = sum(runs_off_bat)) %>%
      filter(runs >= 100) %>% count() %>% ungroup() %>%
      rename(hundreds = n)

    not_out <- final_df %>% filter(grepl(x, season.x)  &
                                     (striker %in% striker_interest |
                                        non_striker %in% striker_interest)) %>%
      filter(ball >= 19.6) %>%
      dplyr::distinct(match_id) %>%
      count() %>%
      rename(not_out = n) %>%
      mutate(striker = striker_interest)#number of times not out

    runs <- df_interest %>%
      summarise(runs = sum(runs_off_bat)) %>%
      mutate(striker = striker_interest)

    matches <- final_df %>% filter(grepl(x, season.x)  &
                                     (striker %in% striker_interest |
                                        non_striker %in% striker_interest)) %>%
      dplyr::distinct(match_id) %>%
      count() %>%
      rename(matches = n) %>%
      mutate(striker = striker_interest)

    row_df <-  reduce(list(matches, fours, sixes, fifties, hundreds, not_out, runs), left_join, by = 'striker')

    row_df <- row_df %>% mutate(avg = round(runs/(matches - not_out), 2)) %>%
      select(-runs)

    bound <- bind_rows(bound, row_df)

    bound <- bound %>% replace(is.na(.), 0)

  }
  y1 <- left_join(bound, y1, by = 'striker')

  y1 <- y1 %>% select(11, 9, 2, 1, everything())

  if (y == "") {
  y1 <- y1 # if there is an empty value, then this changes the rank column
  } else {
 y1 <- y1 %>% mutate(rank = NA) 
  }
  
    if (x >= 2008) {
  y1 <- y1 # if there is an empty value, then this changes the year column
  } else {
 y1 <- y1 %>% mutate(year = NA) 
}
  
  return(y1)
}

```

This function takes year, name of player, and number of entries to return as the input

```{r}
high_scorer(x = 2024, y = , n = 5) # another change in 2025
```

Number of runs of the top scorers across the years
```{r}
# hs2 <- NULL
# for (i in min(ipl_season):max(ipl_season)) {
#   hs <- high_scorer(x = i)
#   hs2 <- rbind(hs2,hs)
# }
# 
# nanoparquet::write_parquet(hs2, "cricket/hs2.parquet")
hs2 <- nanoparquet::read_parquet("cricket/hs2.parquet")

hs2 %>%
  filter(year >= 2011) %>%
  ggplot(aes(x = as.factor(rank),
             y = runs))+
  geom_col()+
  facet_wrap(~year)+
  labs(
    x = "",
    y = "Runs",
    title = "Total runs of the top 10 run scorers over the past seasons")
```



Some players have consistently been in the top-10 over multiple seasons. Though
the trend seems to be decreasing. The top five runscorers repeat many times.
```{r}
(hs_freq <- hs2 %>% filter(year >= 2011) %>%
    count(striker) %>% arrange(desc(n)) %>% head(5))
```


Let's visualise this

```{r}
hs2 %>%
  left_join(hs_freq, by = "striker") %>%
  mutate(is_hs = ifelse(n > 0, '1', '0')) %>%
  filter(year >= 2011) %>%
  ggplot(aes(x = as.factor(rank),
             y = runs,
             fill = is_hs))+
  geom_col()+
  facet_wrap(~year)+
  labs(
    x = "",
    y = "Runs",
    title = "<b style='color:#f8766d'>Top five</b> most runscorers
             perform consistently.")+
  theme(plot.title = element_markdown(),
        legend.position = "none")

```




### Are first innings runs increasing over the years?

We can visualise this by plotting runs in the first innings and matches from the beginning of IPL
```{r Plotly plot}
first_innings_run <- final_df %>%
  mutate(runs = rowSums(select(.,runs_off_bat,extras), na.rm = TRUE)) %>%
  group_by(match_id, innings) %>%
  summarise(total_runs = sum(runs)) %>%
  filter(innings == 1) %>%
  left_join(.,final_df,by = 'match_id') %>%
  distinct(match_id, .keep_all = TRUE) %>%
  ungroup() %>%
  mutate(match_no = row_number()) %>%
  select(match_no, everything())

p2 <- first_innings_run %>%
  ggplot(aes(x = match_no, y = total_runs))+
  geom_point(aes(color = as.factor(season.x))) +
  geom_smooth(method = "lm")+
  guides(color = guide_none())+
  labs(
    x = "Matches",
    y = "Number of runs in first innings",
    title = "Noticeable increase in the runs (Hover for more details)"
  )

ggplotly(p2)

```

## What about boundaries (4s and 6s)
```{r}
boundaries <- final_df %>% 
    group_by(match_id, runs_off_bat) %>% 
    count(runs_off_bat) %>% filter(runs_off_bat == 4 | runs_off_bat == 6 ) %>%
  ungroup() %>% 
  mutate(match_no = row_number()) %>%
  select(match_no, everything())

p5 <- boundaries %>% 
    inner_join(.,first_innings_run, by = 'match_id') %>% 
    ggplot(aes(x = match_no.y, y = n, color = as.factor(runs_off_bat.x))) +
    geom_point()+
    geom_smooth()+
    labs(
        x = "Matches",
        y = "Number of boundaries per match",
        title = "Boundaries and IPL matches",
        col = "4/6"
    )
ggplotly(p5)

```


### Stats
A simple linear model for testing for increase in runs over time
```{r}
lm(total_runs ~ match_no, first_innings_run) %>% summary()
```

